---
title: "Methylation IP vs rest"
author: "almut"
date: "07 October 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Methylation groups
Analyse genes associated with methylation groups

libraries
```{r load libraries, warning=F, message=FALSE}
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(ggpubr)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(here)
library(piano)
```

# Data preprocessing

data
```{r load data}
data_dir <- here("data")
output_dir <- here("output")
figure_dir <- here("output/figures")

#dds data set. gene expression data + patmetadata
load(paste0(data_dir, "/ddsrnaCLL_150218.RData"))

#arrange columns
mutStatus <- data.frame(colData(ddsCLL)) %>% arrange(Methylation) %>% filter(!is.na(Methylation)) %>% mutate("IP" = ifelse(Methylation %in% c("LP", "HP"), 0, 1))
colnames(ddsCLL) <-colData(ddsCLL)$PatID
ddsCLL <- ddsCLL[, mutStatus$PatID]
colData(ddsCLL)$IP <- mutStatus$IP
table(colData(ddsCLL)$IP)
```

Normalize
```{r RNAnorm}
#expression data
ddsCLL <- estimateSizeFactors(ddsCLL)
RNAnorm <- varianceStabilizingTransformation(ddsCLL, blind = T)

```

# Exploratory data analysis

Clustering on the most variable genes already split methylation groups. How do methylation groups affect highly variable genes? 
Can we distinguish all 3 groups?

PCA
```{r pca, fig.height=6, fig.width=6}
#Plot PCA
exprMat <- assay(RNAnorm)

#top 5000 most variant genes
sds <- rowSds(exprMat)
exprMat <- exprMat[order(sds, decreasing = T)[1:150],]


#Calculate PCA
pcaRes <- prcomp(t(exprMat), scale =T)
varExp <- (pcaRes$sdev^2 / sum(pcaRes$sdev^2)) * 100
pcaTab <- data.frame(pcaRes$x[,c(1:10)])
names(varExp) <- colnames(pcaRes$x)

#add background information
pcaTab <- cbind(pcaTab, data.frame(colData(RNAnorm)))


#plot PCA and color samples based on annotations
annocol <- get_palette("jco", 10)
  
p <- ggscatter(pcaTab, x = "PC1", y = "PC2", color = "Methylation", palette = c( annocol[7], annocol[5], annocol[6]),
  ylab = sprintf("PC2 (%2.1f%%)",varExp[2]), xlab = sprintf("PC1 (%2.1f%%)",varExp[1]), legend = "right", main = "PCA Methylation groups") + coord_fixed()

p

#ggsave(file=paste0(figure_dir, "/pca_Meth_IP_all_top150.svg"), plot=p, width=5, height=5)
```

# Gene expression

##Differential expression analysis

deseq2
```{r deseq2, eval = FALSE}

###Deseq
ddsCLL <- estimateSizeFactors(ddsCLL)

# deseq2 function
diff <- function(cond){
  ddsCLL_new <- ddsCLL[,!is.na(colData(ddsCLL)[,cond])]
  design(ddsCLL_new) <- as.formula(paste("~ ", paste(cond)))
  rnaRaw <- DESeq(ddsCLL_new, betaPrior = FALSE)
  res <- results(rnaRaw)
  resOrdered <- res[order(res$pvalue),]
}

#diff_meth <- diff("IP")

#saveRDS(diff_meth, file= paste0(output_dir,"/diff_meth_IP_vs_all.rds"))


```

Filter differentially expressed genes
```{r diff genes}

diff_meth <- readRDS(paste0(output_dir,"/diff_meth_IP_vs_all.rds"))

dataTab <- data.frame(diff_meth)
dataTab$ID <- rownames(dataTab)

#filter using pvalues
dataTab <- dataTab %>%
    arrange(padj) %>%
    mutate(Symbol = rowData(ddsCLL[ID,])$symbol)
dataTab <- dataTab[!duplicated(dataTab$Symbol),]
dataTab <- dataTab[!is.na(dataTab$Symbol),]
rownames(dataTab) <- dataTab$ID
write.csv(dataTab, file=paste0(output_dir, "/diff_genes/meth_IP_vs_all_diffGenes.csv"))

```


## Expression heatmap

```{r complex heatmap, fig.width= 30, fig.height=45}
#arrange columns
mutStatus <- data.frame(colData(ddsCLL)) %>% arrange(Methylation)
colnames(ddsCLL) <-colData(ddsCLL)$PatID
ddsCLL <- ddsCLL[, mutStatus$PatID]


#Differentially expressed genes
genes <- dataTab %>% filter(padj <= 0.01, abs(stat) > 6)  
exprMat <- assay(RNAnorm)
exprMat<- exprMat[genes$ID,]

#scale gene expression
colnames(exprMat) <- colData(ddsCLL)$PatID
exprMat.new <- log2(exprMat)
exprMat.new <- t(scale(t(exprMat.new)))
exprMat.new[exprMat.new > 4] <- 4
exprMat.new[exprMat.new < -4] <- -4

#colors
colors <- colorRampPalette(rev( brewer.pal(10,"RdBu")) )(20)
annocol <- get_palette("jco", 10)
annocolor <- list(Methylation = c("IP" = annocol[5], "LP" = annocol[6], "HP" =  annocol[7]), IGHV = c("M" = annocol[1], "U" = annocol[2])) 

#Annotation
feature <- as.data.frame(colData(ddsCLL)[,c("Methylation", "IGHV")]) 
colnames(feature) <- c("Methylation", "IGHV") 

#gene symbol as rownames
rownames(exprMat.new) <- rowData(RNAnorm[rownames(exprMat),])$symbol

ha_col <- HeatmapAnnotation(df = feature, col = annocolor, annotation_height = unit(c(rep(1.3, 2)), "cm"), annotation_legend_param = list(title_gp = gpar(fontsize = 40), labels_gp = gpar(fontsize = 35),  grid_height = unit(1.9, "cm"), grid_width = unit(1.9, "cm")))


h1 <- Heatmap(exprMat.new ,                                                    
              km = 2,
              cluster_columns = F,
              clustering_distance_rows = "pearson",
              clustering_method_rows = "ward.D2",
              column_title ="Gene signature methylation groups ",                      
              col = colors,
              column_title_gp = gpar(fontsize = 50, fontface = "bold"), 
              heatmap_legend_param = list(title = "expr", 
                                          title_gp = gpar(fontsize = 40), 
                                          grid_height = unit(1.9, "cm"), 
                                          grid_width = unit(1.9, "cm"), 
                                          gap = unit(2, "cm"), 
                                          labels_gp = gpar(fontsize = 35)), 
              show_row_dend = FALSE, 
              show_column_names = FALSE , 
              show_row_names = FALSE, 
              row_names_gp = gpar(fontsize = 21),
              top_annotation = ha_col)



#Annotate top 50 genes
sub_names <- genes[1:50,"Symbol"]
sub_names <- sub_names[-which(sub_names %in% "")]
geneIDs <- which(rownames(exprMat.new) %in% sub_names)
labels <- rownames(exprMat.new)[geneIDs]
ha_genes <- rowAnnotation(link = row_anno_link(at = geneIDs, labels = labels, labels_gp = gpar(fontsize = 35)), width = unit(9, "cm"))


#svg(filename=paste0(figure_dir, "/gene_expr_Methylation_IP_all.svg"), width=30, height=35)
#pdf(file=paste0(figure_dir,"/gene_expr_Methylation_IP_all.pdf"), width=30, height=45)
p1 <- draw( h1 + ha_genes)
#dev.off()

saveRDS(p1, file = paste0(output_dir, "/figures/r_objects/Methylation_IP/meth_ip_heatmap.rds"))


```


# Gene and Sample specific expression - top genes
```{r single gene counts, message=FALSE}
#function to create stripchart plots for specific genes
gene_count <- function(gene_nam){
  geneEnsID <- rownames(ddsCLL)[which(rowData(ddsCLL)$symbol %in% gene_nam)]
  geneNum <- exprMat[geneEnsID,]
  mutPat <- as.data.frame(colData(ddsCLL)[, c("Methylation")])
  colnames(mutPat) <- c("genotype")
  geneDat <- cbind(mutPat, geneNum)
  colnames(geneDat) <- c("genotype", "counts")
  
  p <- ggstripchart(geneDat, x = "genotype", y = "counts",
          color = "genotype",
          palette = "jco",
          add = "mean_sd",
          title = paste(gene_nam),
          ylab = "normalized counts")
 # ggsave(file=paste0(figure_dir, "/methylation_IP_all/genetic_interaction_", gene_nam, ".svg"), plot=p, width=6, height=5)
  saveRDS(p, file = paste0(output_dir, "/figures/r_objects/Methylation_IP/de_genes/", gene_nam, ".rds"))
  p
}

geneList <- sub_names[1:30]
lapply(geneList, gene_count)
gene_count("SOX11")

```

# Gene set enrichment
```{r gsea, fig.width=16, fig.height=7}

variant <- "IP"
gmtFile <- loadGSC(paste0(data_dir,"/c2.cp.kegg.v6.0.symbols.gmt"), type="gmt")

diff_res <- dataTab
diff_res  <- diff_res[-which(diff_res$Symbol %in% c("", NA)),]
#get genes and pvalues 
geneNam <- diff_res$Symbol 
pVal <- diff_res$padj
logFold <- diff_res$log2FoldChange
stat <- diff_res$stat
gsTab <- data.frame(gene = geneNam, stat = stat)
gsaTab <- data.frame(row.names = gsTab$gene, stat = gsTab$stat)
res <- runGSA(geneLevelStats = gsaTab,
                      geneSetStat = "gsea",
                     adjMethod = "fdr", gsc=gmtFile,
                     signifMethod = "geneSampling",
                      nPerm = 50000,
                      gsSizeLim=c(1, Inf))
Res_up <- arrange(GSAsummaryTable(res), `p adj (dist.dir.up)`)                 
Res_dn <- arrange(GSAsummaryTable(res), `p adj (dist.dir.dn)`)
  

#Plot
resPlot <- Res_dn[, c(1:3,7,8,9)]
colnames(resPlot) <- c("pathway", "gene_number", "stat", "p.adj","genes_up" , "genes_dn")


enrichPlot <- resPlot %>% filter(p.adj < 0.1) %>% mutate(log10Padj = -log10(p.adj)) #%>% mutate(genes = ifelse(gene_number > 5, ">5", "<=5"))
enrichPlot$log10Padj[which(enrichPlot$log10Padj == Inf)] <- 5

p <- ggbarplot(enrichPlot, x = "pathway", y = "log10Padj",
          fill = "gene_number",         
          color = "white",           
          palette =  "gsea",            
          sort.val = "asc",          
          sort.by.groups = FALSE,     
          ylab = "-log10(padj)",
          legend.title = "#diff.genes",
          rotate = TRUE,
          font.x = 20, font.y = 20, font.legend = 20, legend = "right", 
          title = "Methylation groups - Kegg",
          ggtheme = theme_pubr()) +
  font("xy.text", size = 16) + 
  font("title", size = 20, face = "bold")
  
#ggsave(file=paste0(figure_dir, "/GSEA_Meth_IP_vs_all_Kegg.svg"), plot=p, width=14, height=7)

p


saveRDS(p, file = paste0(output_dir, "/figures/r_objects/Methylation_IP/meth_ip_enrichment.rds"))
```

