%%%%%%%%%%%%%%%%%%% Title, Table of Contents etc. %%%%%%%%%%%%%%%%%%%%%%%

\begin{titlepage}
	\centering
	{\normalsize Ruprecht-Karls-Universität Heidelberg \par}
	\vspace{0.25cm}
	{\normalsize Fakultät für Biowissenschaften \par}
	\vspace{0.25cm}
	{\normalsize Masterstudiengang Molekulare Biotechnologie \par}
	\vspace{3.5cm}
	{\LARGE \bfseries Systematic Expression Profiling of Chronic Lymphatic Leukemia Transcriptomes \par}
	\vspace{5cm}
	{\normalsize Masterarbeit \par}
	\vspace{4cm}
	{\normalsize\textbf{ Von Almut Lütge} \par}
	\vspace{0.25cm}
	{\normalsize\textbf{Aus Hildesheim} \par}
	\vspace{0.5cm}
	{\normalsize\textbf{Abgabetermin Dezember 2017}}
\end{titlepage}


\cleardoublepage

\begin{titlepage}

Die vorliegende Masterarbeit wurde am Deutschen Krebsforschungszentrum Heidelberg in der Abteilung für Computational Oncology in der Zeit von 08/06/2017 bis 08/12/2017 angefertigt.\par
\vspace{2.5cm}
Gutachter der Arbeit: \hspace*{12mm}	 Dr. Matthias Schlesner\par
					\hspace*{54mm}	Computational Oncology\par
					\hspace*{54mm}	Deutsches Krebsforschungszentrum Heidelberg\par
\vspace{1cm}
Zweitgutachter der Arbeit: \noindent\hspace*{3mm}	
Dr. Wolfgang Huber\par
\noindent\hspace*{56mm}Structural and Computational Biology\par
\noindent\hspace*{56mm}European Molecular Biology Laboratory \par
\vspace{2.5cm}
\hrule
\vspace{1.5cm}
Ich erkläre hiermit ehrenwörtlich, dass:
\begin{enumerate}
\item ich die vorliegende Masterarbeit selbständig unter Anleitung verfasst und keine anderen als die angegebenen Quellen und Hilfsmittel benutzt habe;
\item die Übernahme wörtlicher Zitate aus der Literatur/Internet sowie die Verwendung der Gedanken anderer Autoren an den entsprechenden Stellen innerhalb der Arbeit gekennzeichnet wurde;
\item ich meine Masterarbeit bei keiner anderen Prüfung vorgelegt habe.
\end{enumerate}
\vspace{0.25cm}
Ich bin mir bewusst, dass eine falsche Erklärung rechtliche Folgen haben wird. \par

\vspace{3.5cm}


Ort, Datum \hspace*{40mm}			Unterschrift\par


\end{titlepage}
\cleardoublepage
\onehalfspacing

\section*{Acknowledgements}

First of all, I want to thank Dr.~Junyan Lu and Dr.~Wolfgang Huber, who supervised this work with all their expertise and inspiring comments and feedback, but also invested a lot of time and patience to support me. I also want to thank for the chance to work on this project and even more the encouragement to learn so many new things and actively participate in all the exciting conferences, courses and retreats.\\
I want to thank Dr.~Matthias Schlesner and Dr.~Thorsten Zenz for their feedback and additional expertise. It was a real support. 
Many thanks to all members of the Huber group for the wonderful time, the support and all the chats at lunch or coffee breaks. I really enjoyed working with them. \\
My special thanks are to my twinsister Mechthild for her graceful support in every situation as well as to my sisters Barbara, Rosa and my parents for always being my backup.   

